---
title: "B-splines and P-splines"
author: "Martin Liermann"
format: html
editor: visual
---

B-splines or (basis splines) are piece wise polynomial functions that can be expressed as a linear combination of basis functions, $B_{k,d}$'s.

$$
f(x) = \sum_{k=1}^N{\alpha_kB_{k,d}(x)}
$$

The $B_{k,d}$ functions can be any order, $d$, but are typically order 4, which corresponds to a 3rd degree polynomials (cubic, $d=4$). The basis function $B_{k,d}(x)$ is non-zero only between the knots $u_k$ and $u_{k+d}$ where the knots $u_1, u_2,…u_N$ are sequential values that subdivide the range of $x$ over which you would like to estimate $y$. Between $u_k$ and $u_{k+d}$, the $B_{k,d}$'s can be defined recursively as:

$$
B_{k,1}(x) = 1 \ \ \text {if} \ \ x \in (u_k,u_{k+1}] \ \ \text {else} \ \ 0
$$

$$
B_{k,d}(x) = \frac {x-u_{k}}{u_{k+d-1}-u_{k}} B_{k,d-1}(x) + \frac {u_{k+d}-x}{u_{k+d}-u_{k+1}} B_{k+1,d-1}(x) \\
$$

These are known as the Cox-deBoor Equations. Notice that for a given set of $x$'s and specific order, $d$,you end up with a matrix, $B$ , with a row for each $x$ and a column for each knot, $u$.

Here I implement the Cox-deBoor recursive relationship in R and plot the order 4 basis functions. Here I have explicitly reproduced the recursive equation above. The Cox-deBoor algorithm is a more efficient implementation.

```{r}
u <- (0:10)*10
uN <- length(u)
x <- seq(1,99,by=0.1)
xN <- length(x)
B <- array(NA,dim=c(xN,uN,4))
for(d in 1:4){ # iterate over orders, d
  for(k in 1:(uN-d)){ # iterate over knots, u[k]
    for(j in 1:xN){ # iterate over x values, x[j]
      if(d==1){
        B[j,k,d] <- ifelse(x[j]<u[k] | x[j]>=u[k+1],0,1)
      }else{
        B[j,k,d] <- (x[j]-u[k])/(u[k+d-1]-u[k])   * B[j,k,d-1] +
                    (u[k+d]-x[j])/(u[k+d]-u[k+1]) * B[j,k+1,d-1]
      }
    }
  }
}

# plot the basis functions for order 4 (d=4, cubic)
ymax <- max(B,na.rm=TRUE)
plot(1,1,xlim=range(x),ylim=c(0,ymax),type="n",xlab="x",ylab="",bty="l")
for(j in 1:uN){
  lines(x,B[,j,4],col=j)
}
```

The order 1 basis function for each knot, $k$, is 1 between $u_k$ and $u_{k+1}$ and 0 everywhere else.

```{r}
ymax <- max(B,na.rm=TRUE)
plot(1,1,xlim=range(x),ylim=c(0,ymax),type="n",xlab="x",ylab="",bty="l")
for(j in 1:uN){
  lines(x,B[,j,1],col=j)
}
```

The order 2 basis function is the triangular function (positive between $u_k$ and $u_{k+2}$).

```{r}
ymax <- max(B,na.rm=TRUE)
plot(1,1,xlim=range(x),ylim=c(0,ymax),type="n",xlab="x",ylab="",bty="l")
for(j in 1:uN){
  lines(x,B[,j,2],col=j)
}
```

The P-spline is a B-spline with a penalty.

$$
||y-f(x)||^2  + \lambda ||\alpha_i - \alpha_{i-1}||^2 = ||y-B\alpha||^2  + \lambda ||D\alpha||^2
$$

Here, the $D$ matrix is a first order differencing matrix.

```{r}
D <- matrix(0,ncol=4,nrow=4)
D[1,1] <- 1
for(i in 2:4){
  D[i,i]<-1
  D[i,i-1] <- -1
}
knitr::kable(D,format.args=list(trim=FALSE,width=4))
```

Using some linear algebra, you can find the $\alpha$ that minimizes the sum of the squared differences between the $y$ and $B\alpha$, given a penalty function.

$$
\hat {\alpha} = (B'B+\lambda D'D)^{-1}B'y
$$

So if I simulate some data, I can fit using different $\lambda$'s.

```{r}
x <- 1:99
y <- 0.002*(x-50)^2+0.01*x + rnorm(99,0,1)
plot(x,y,bty="l",pch=16)
```

Now I create a basis with knots that extend on either side of the data.

```{r}
u <- (0:40)*5-50
uN <- length(u)
x <- 1:99
xN <- length(x)
B <- array(NA,dim=c(xN,uN,4))
for(d in 1:4){ # orders
  for(k in 1:(uN-d)){ # iterate over knots
    for(j in 1:xN){ # iterate over x values
      if(d==1){
        B[j,k,d] <- ifelse(x[j]<u[k] | x[j]>=u[k+1],0,1)
      }else{
        B[j,k,d] <- (x[j]-u[k])/(u[k+d-1]-u[k])   * B[j,k,d-1] +
                    (u[k+d]-x[j])/(u[k+d]-u[k+1]) * B[j,k+1,d-1]
      }
    }
  }
}

D <- matrix(0,ncol=uN-4,nrow=uN-4)
D[1,1] <- 1
for(i in 2:(uN-4)){
  D[i,i]<-1
  D[i,i-1] <- -1
}
BB <- B[,1:(uN-4),4]
```

And plot the data and P-spline with different penalties. It looks like a penalty of around 3 works best for this problem.

```{r}
plot(x,y,bty="l",pch=16,col=rgb(0,0,0,0.2))
pp <- c(1,3,10,100)
for(i in 1:4){
  zz <- solve(t(BB)%*%BB + pp[i]*t(D)%*%D) %*% t(BB) %*% y
  yPred <- BB %*% zz
  lines(x,yPred,col=i,lwd=2)
}
legend("bottomright",pch=16,col=1:4,legend=paste(pp))
```

Next I could:

-   Look at how best to expand on either side, natural splines, etc..
-   Use cross-validation to find the penalty
